Co-evolution of lexical meaning \& pragmatic use

### 2018-01-11 paper/cogsci-submission2.tex
revisions semi-draft

\definecolor{Red}{RGB}{178,34,34}
\newcommand{\mf}[1]{\textcolor{Red}{[MF: #1]}} 
\newcommand{\tb}[1]{\textcolor[rgb]{.8,.33,.0}{[TB: #1]}}% prints in orange

@@

\begin{abstract}
  According to standard linguistic theory, the meaning of an utterance is the product of
  conventional semantic meaning and general pragmatic rules on language use. To investigate how
  cultural evolution of language plays out under this picture of the semantics-pragmatics
  interface, we present a game theoretic model of the competition between types of language
  users, each endowed with a selection of lexical representations and a particular pragmatic
  disposition to act on them. Our model traces two evolutionary forces and their interaction:
  (i) pressure towards communicative efficiency and (ii) learning biases during the transfer of
  linguistic knowledge. We illustrate the model based on a case study on scalar
  implicatures. In this case study learning biases that favor simple semantic representations
  can foster the evolution of more sophisticated pragmatic reasoning types and so prevent the
  lexicalization of scalar implicatures.
\end{abstract}

\section{Introduction}\label{sec:introduction}
What is conveyed usually goes beyond what is said. A request for a blanket can be politely
veiled by uttering ``I'm cold;'' temporal succession of events can be communicated by the order
in which conjuncts appear as in ``I traveled to Paris and got married;'' an invitation can be
declined by saying ``I have to work.'' An influential explanation of the relation between the
literal meaning of expressions and what they may convey in context is due to
\citet{grice:1975}, who characterizes pragmatic use and interpretation as a process of mutual
reasoning about rational language use. For instance, under the assumption that the speaker is
cooperative and relevant, ``I have to work'' may be interpreted as providing a reason why the
speaker will not be able to accept an invitation, going beyond its literal meaning. Some of
these enrichments are rather \emph{ad hoc}. Others show striking regularities, such as the use
of ability questions for polite requests (``Could you please \dots?''), or certain enrichments
of lexical meanings such as \emph{and} to convey \emph{and then}.

A particularly productive and well studied class of systematic pragmatic enrichments are scalar
implicatures
\citep{horn:1984,Hirschberg1985:A-Theory-of-Sca,LevinsonPragmatics1983,Geurts2010:Quantity-Implic}. Usually,
the utterance of a sentence like ``I own some of Johnny Cash's albums'' will be taken to mean
that the speaker does not own all of them. This is because, if the speaker instead owned them all, she
could have used the word \emph{all} instead of \emph{some} in his utterance, thereby making a
more informative statement. Scalar implicatures, especially the inference from \emph{some} to
\emph{some but not all}, have been studied extensively, both theoretically
\citep[e.g.,][]{Sauerland2004:Scalar-Implicat,ChierchiaFox2008:The-Grammatical,Rooyvan-RooijJagerde-Jager2012:Explaining-Quan}
as well as experimentally
\citep[e.g.,][]{BottNoveck2004:Some-Utterances,huang+snedeker:2009,GrodnerKlein2010:Some-and-Possib,GoodmanStuhlmuller2013:Knowledge-and-I,DegenTanenhaus2012:Processing-Scal}. While
there is much dispute in this domain about many details, a position endorsed by a clear
majority is that a scalar item like \emph{some} is underspecified to mean \emph{some and maybe
  all} and that the enrichment to \emph{some but not all} is part of some regular process with
roots in pragmatics.

If this majority view is correct, the question arises how such a division of labor between
semantics and pragmatics could have evolved, why it would be so pervasive across natural
languages, and why it is that some expressions systematically draw from it while others semantically conventionalize. Models of language evolution abound. There are simulation-based models studying
populations of communicating agents
\citep{Hurford1989:Biological-Evol,Steels1995:A-Self-Organizi,LenaertsJansen2005:The-Evolutionar,SteelsBelpaeme2005:Coordinating-Pe,BaronchelliPuglisi2008:Cultural-route-,steels:2011,SpikeStadler2016:Minimal-Require}
and there are mathematical models of language evolution, mostly coming from game theory
\citep{lewis:1969,Warneryd1993:Cheap-Talk-Coor,BlumeKim1993:Evolutionary-St,nowak+krakauer:1999,Huttegger2007:Evolution-and-t,Skyrms2010:Signals}. Much
of this work has focused on explaining basic properties such as compositionality and
combinatoriality
\citep[e.g.,][]{Batali1998:Computational-S,nowak+krakauer:1999,nowak+etal:2000,KirbyHurford2002:The-Emergence-o,kirby:2002,SmithKirby2003:Iterated-Learni,Gong2007:Language-Evolut,kirby+etal:2015,verhoef+etal:2014,Franke2015:Proto-Syntax},
but little attention has been paid to the interaction between conventional meaning and
pragmatic use. What is more, many mathematical models explain evolved meaning as a regularity
in the overt behavior of agents, abstracting from complex interactions between semantic
representations and pragmatic use. In contrast, we will here look at language users with a
richer cognitive make-up.

We spell out a model of the co-evolution of conventional meaning and pragmatic reasoning.
The objects of replication and selection are pairs consisting of a set of lexical
meanings and a manner of pragmatic behavior. The latter are represented using probabilistic
models of pragmatic language use
\citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}. Replication
and selection are described by the \emph{replicator mutator dynamic}, a general and established
model of evolutionary change in large and homogeneous populations
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. This
approach allows us to study the interaction between (i) pressure towards communicative
efficiency and (ii) infidelity in the transmission of linguistic knowledge, caused by factors
such as inductive learning biases and sparse learning data. Considering transmission of
linguistic knowledge is important because neither semantic meanings nor pragmatic usage
patterns are directly observable. Instead, language learners have to infer these unobservables
from the observable behavior in which they result. We formalize this process as a form of
Bayesian inference. Our approach thereby contains as a special case a well-understood model of
iterated Bayesian learning \citep{griffiths+kalish:2007}, but combines it with functional
selection, here formalized as the most versatile dynamic from evolutionary game theory, the
replicator dynamic \citep{TaylorJonker1978:Evolutionary-St}. Section~\ref{sec:model} introduces
this model.

Section~\ref{sec:si-case-study} applies this model to a case study on scalar implicatures. We
discuss a setting in which the majority view of underspecified lexical meanings and systematic pragmatic
enrichments emerges if selection and transmission infidelity are combined. In particular, we
show that inductive learning biases of Bayesian learners that favor simpler lexical meanings
can prevent the lexicalization of scalar inferences and lead to the emergence of Gricean-like
pragmatic reasoning types. Results of this case study are critically assessed in the light of
the assumptions that feed our model in Section~\ref{sec:discussion}.

@@

\subsection{Communicative efficiency and learnability}

@@

The idea that language is an adaptation to serve a communicative function is fundamental to
many synchronic and diachronic analyses at least since \citeposs{zipf:1949} explanation of word
frequency rankings as a result of competing hearer and speaker preferences \citep[e.g.,
in][]{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007,
  piantadosi:2014,kirby+etal:2015}. If processes of selection, such as conditional imitation or
reinforcement, favor more communicatively efficient types of behavior, languages are driven
towards semantic expressivity \citep[e.g.,][]{nowak+krakauer:1999,Skyrms2010:Signals}. But
pressure towards communicative efficiency is not the only force that shapes
language. Learnability is another, as natural languages need to be learnable to survive their
faithful transmission across generations. Furthermore, even small learning biases implicit in acquisition
can build up and have quite striking effects on an evolving language in a process of iterated
learning
\citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,kirby+etal:2014}.

While natural languages are pressured for both communicative efficiency and learnability, these forces may
pull in opposite directions. Their opposition becomes particularly clear when considering the
extreme (cf. \citealt{kemp+regier:2012,kirby+etal:2015}). A language consisting of a single
form-meaning association is easy to learn but lacking in expressivity. Conversely, a language
that lexicalizes a distinct form for a large number of different meanings is highly expressive
but challenging to acquire.

\subsection{The replicator mutator dynamic}

An elegant formal approach to capture the interaction between communicative efficiency and learnability is
the \emph{replicator mutator dynamic}
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. In
its simplest, discrete-time formulation, the RMD defines the frequency $x'_i$ of each type $i$
in a population at the next time step as a function of: (i) the frequency $x_i$ of each type
$i$ before the update step, (ii) the fitness $f_i$ of each type $i$ before the update, and
(iii) the probability $Q_{ji}$ that an agent who wants to imitate, adopt, or learn the type of
an agent with type $j$ ends up acquiring type $i$:
\begin{align}
  \label{eq:RMD_discrete}
  x'_i = \sum_j Q_{ji} \frac{x_jf_j}{\sum_k x_k f_k}\,.
\end{align}
The RMD consists of two components: fitness-based selection and transmission biases. This
becomes most transparent when we consider an equivalent formulation in terms of a step-wise
application of the discrete-time replicator dynamic \citep{TaylorJonker1978:Evolutionary-St} on the initial population vector $\vec{x}$
and its subsequent multiplication with a mutation matrix $Q$:
\begin{align}
  \label{eq:RMD_discrete_recast}
  x'_i & = (\text{M}(\text{RD}(\vec{x})))_i\,,
\end{align}
where
\begin{align*}
      \left ( \text{RD}(\vec{x}) \right )_i 
         = \frac{x_i f_i}{\sum_k x_k f_k}
 \ \ \ \ \text{and} \ \ \ \ 
  (\text{M}(\vec{x}))_i = (\vec{x} \cdot Q)_i = \left ( \sum_j
          x_j Q_{ji} \right)_i\,.
\end{align*}
If the transmission matrix $Q$ is trivial in the sense that $Q_{ji}=1$ whenever $j=i$, the
dynamic reduces to the replicator dynamic. The replicator dynamic is a model of fitness-based
selection in which the relative frequency of type $i$ will increase with a gradient
proportional to its average fitness in the population. This dynamic is popular and
versatile because it can be derived from many abstract processes of biological and cultural
transmission and selection \citep[for overview and several derivations
see][]{Sandholm2010:Population-Game}, including conditional imitation
\citep[e.g.,][]{Helbing1996:A-Stochastic-Be,Schlag1998:Why-Imitate-and} or reinforcement
learning \citep[e.g.,][]{BorgersSarin997:Learning-Throug,Beggs2005:On-the-Converge}. If fitness
$f_i$ is the same for all types $i$, the replicator step is the identity map
$ \left ( \text{RD}(\vec{x}) \right )_i = x_i$ and the dynamic reduces to a process of
iteration of the transmission bias encoded in $Q$. In this way, the process in
(\ref{eq:RMD_discrete}), equivalently (\ref{eq:RMD_discrete_recast}), contains a model of
iterated learning \citep{griffiths+kalish:2007}.

@@

\begin{example}
  Consider a simple and abstract coordination game. Agents are of two types: positive or negative. If
  agents of different type play with each other, they obtain a payoff of 0. If negative meets
  negative, each receives a payoff of 1. If positive meets positive, they get a payoff of 2. A
  population state is completely characterized by the proportion $x$ of negatives. The fitness
  of negatives in population state $x$ is $f_n(x) = x$, that of positives is $f_p(x) = 2-2x$.
  The average fitness is $\Phi(x) = x f_n(x) + (1-x) f_p(x) = 3x^2 - 4x + 2$. Without mutation,
  the replicator dynamic will then update $x$ to $RD(x) = \nicefrac{f(x)x}{\Phi(x)} = \nicefrac{x^2}{\Phi(x)}$. The update
  function $RD(x)$ of the replicator step is plotted in Fig.~\ref{fig:Updates_RMS} as the
  blue line. Rest points, for which $RD(x)=x$, are at $x=0$, $x=1$ and $x= \nicefrac{2}{3}$.
  The former are attractors as nearby points converge to them. Points near $x=\nicefrac{2}{3}$
  move towards 0 or 1. This is schematically pictured in the topmost phase portrait in
  Fig.~\ref{fig:Phase_RD}.

  Adding mutation changes the dynamic and its rest points. Let us assume that $Q_{ji} = .9$
  when $j=i$. The update effect of mutation alone is $M(x) = .9 x + .1 (1-x) = .8x + .1$ and is
  plotted as the linear green line in Fig.~\ref{fig:Updates_RMS}. It has only one stable rest
  point at $x = 0.5$ (see Fig.~\ref{fig:Phase_RD}). If we first take the replicator step and
  then the mutation step in sequence, we obtain the replicator mutator dynamic
  $RMD(x) = M(RD(x)) = \nicefrac{.9x^2 - .2x +.2}{3x^2 - 4x + 2}$, which is plotted in red in
  Fig.~\ref{fig:Updates_RMS}. The rest points are at $x=.121$, $x=.903$ and $x=.609$. The
  former two are attractors (see Fig.~\ref{fig:Phase_RD}).
\end{example}

@@

Moving beyond an abstract example, our goal is to apply the RMD to investigate the co-evolution
of lexical representations and pragmatic behavior. To do so, we need to fix three things: (i) what the
relevant types are, (ii) how fitness derives from communicative success and (iii) how the
mutation matrix $Q$ is computed. These issues are addressed, one by one, in the following.

\subsubsection{Types: Lexica and pragmatic strategies}
\label{sec:languages+use}

Types are what evolution operates on. They define an agent's fitness, usually through a payoff
accrued in single interactions with other agents. Often types can be identified as the possible
acts in a game, e.g., either cooperating or defecting in a prisoner's dilemma. In other cases,
they may be thought of as general properties of an agent that influences her fitness, such as
being positive or negative in our previous example (whatever that means). For our present
purposes, types are identified more concretely by specific assumptions about their cognitive
make-up. Since we are interested in the evolutionary competition between different lexical
representations and ways of using them in communication, a type is here defined as a pair consisting
of a lexicon and a pragmatic strategy of language use.

Agents play signaling games, in which the speaker wants to communicate a world state $s$ with a
message $m$ to a hearer who receives $m$ but does not know $s$
\citep[e.g.][]{lewis:1969,Skyrms2010:Signals}. A lexicon associates each message with a
(possibly fuzzy) set of states. A pragmatic behavior specifies a probabilistic speaker
rule (a probabilistic choice of message for each state) and a probabilistic hearer rule (a
probabilistic choice of state for each message) given a lexicon. There are many ways of making these general
notions more concrete. Here is what we will assume in the remainder of this paper.

@@

The following examples illustrate these behaviors. A literal interpreter with lexicon $\Lbound$ assigns $\ssome$

@@

Recall that types are a combination of a lexicon and a manner of language use. We analyze the
model's predictions in populations of types with one of the two behaviors introduced earlier;
literal or pragmatic. The former correspond to level $0$ reasoners and the latter to level
$1$. Accordingly, we consider a total of $432$ types. Six are variants of pragmatic language
users with $\Llack$-like lexica. We refer to these as \emph{target types}, because they
represent lexica and language use that conform to the majority view of scalar
implicatures. Twelve types are either literal or pragmatic speakers with lexica of the
$\Lbound$ kind. We refer to these as \emph{competitor types}, because they are expected to be
the target types' main contenders in evolutionary competition. Finally, note that while
different types may lexicalize the same representations, they may nevertheless map different
states to different overt messages. More informally, they speak different languages that
lexicalize the same concepts. Consequently, more often than not different variants of the same
kind fail to understand each other.

@@

>>>>

\definecolor{Red}{RGB}{178,34,34}
\definecolor{Steelblue}{RGB}{70,130,180}
\newcommand{\mf}[1]{\textcolor{Red}{[MF: #1]}} 
\newcommand{\tb}[1]{\textcolor[rgb]{.8,.33,.0}{[TB: #1]}}% prints in orange
\newcommand{\rvv}[1]{\textcolor{Steelblue}{#1}}% prints in steel blue
\definecolor{indianred}{RGB}{205,92,92}
\definecolor{mygray}{cmyk}{0.35,0.35,0.35,0.35}
\definecolor{goldish}{RGB}{118, 104, 154}

@@

\begin{abstract}
 According to standard linguistic theory, the meaning of an utterance is the product of
  conventional semantic meaning and general pragmatic rules on language use. To investigate how
  cultural evolution of language plays out under this picture of the semantics-pragmatics
  interface, we present a game-theoretic model of the competition between types of language
  users, each endowed with a selection of lexical representations and a particular pragmatic
  disposition to act on them. Our model traces two evolutionary forces and their interaction:
  (i) pressure towards communicative efficiency and (ii) learning biases during the transfer of
  linguistic knowledge. We illustrate the model based on a case study on scalar
  implicatures. \rvv{The relationship between underspecified semantics and pragmatic inference that this study suggests is one of co-evolution. The maintenance of such semantics depends on pragmatic reasoning for their enrichment. In the opposite direction, complex semantics that can do away with mutual reasoning do not necessarily foster pragmatic reasoning and may even be encumbered by it.}
\end{abstract}

\section{Introduction}\label{sec:introduction}
\hl{2. This is true. We are grateful that this was pointed out, as our previous rendering could lead to confusion, e.g., as to why $L_{\text{all}}$, being the {\em a priori} more likely type to be inferred, did not come out strongly even when considering learnability alone. Throughout the article, we now stress more that what matters is the posterior -- and discuss the role of the prior mainly when contrasting targets and competitor, who crucially differ mostly in this respect.}

\hl{3. These questions tie in with Reviewer $1$'s request for a clearer exposition and the explanation of why we speak of co-evolution in this case. In a nutshell, pressure for learnability alone does not put the types found in a population in competition. Instead, the population comes to be inhabited by types that are faithfully transmitted -- of which there might be many (e.g., all variants of a kind are equally learnable). What fitness-relative selection adds to this process is the competition that learnability alone lacks, independent of the type space we look at. Now, in the case of our type space we have a slight (modulo $\lambda$) communicative disadvantage for targets  {\em when in direct competition with competitors}, for instance. However, the fact that they are inferred more faithfully by na\"ive learners leads (i) to the existence of less competitors, counteracting this functional disadvantage, as well as (ii) to the existence of less variants of a kind (contra learning-alone).}

What is conveyed usually goes beyond what is said. A request for a blanket can be politely
veiled by uttering ``I'm cold;'' a temporal succession of events can be communicated by the order
in which conjuncts appear as in ``I traveled to Paris and got married;'' an invitation can be
declined by saying ``I have to work.'' An influential explanation of the relation between the
literal meaning of expressions and what they may convey in context is due to
\citet{grice:1975}, who characterizes pragmatic use and interpretation as a process of mutual
reasoning about rational language use. For instance, under the assumption that the speaker is
cooperative and relevant, ``I have to work'' may be interpreted as providing a reason why the
speaker will not be able to accept an invitation, going beyond its literal meaning. Some of
these enrichments are rather \emph{ad hoc}. Others show striking regularities, such as the use
of ability questions for polite requests (``Could you please \dots?''), or certain enrichments
of lexical meanings such as \emph{and} to convey \emph{and then}.

A particularly productive and well studied class of systematic pragmatic enrichments are scalar
implicatures
\citep{horn:1984,Hirschberg1985:A-Theory-of-Sca,LevinsonPragmatics1983,Geurts2010:Quantity-Implic}. Usually,
the utterance of a sentence like ``I own some of Johnny Cash's albums'' will be taken to mean
that the speaker does not own all of them. This is because, if the speaker instead owned them all, she
could have used the word \emph{all} instead of \emph{some} in his utterance, thereby making a
more informative statement. Scalar implicatures, especially the inference from \emph{some} to
\emph{some but not all}, have been studied extensively, both theoretically
\citep[e.g.,][]{Sauerland2004:Scalar-Implicat,ChierchiaFox2008:The-Grammatical,Rooyvan-RooijJagerde-Jager2012:Explaining-Quan}
as well as experimentally
\citep[e.g.,][]{BottNoveck2004:Some-Utterances,huang+snedeker:2009,GrodnerKlein2010:Some-and-Possib,GoodmanStuhlmuller2013:Knowledge-and-I,DegenTanenhaus2012:Processing-Scal}. This makes them particularly suitable candidates for the study of the evolution of regular pragmatic inferences.  While there is much dispute in this domain about many details, a position endorsed by a clear majority is that a scalar item like \emph{some} is underspecified to mean \emph{some and maybe all} and that the enrichment to \emph{some but not all} is part of some regular process with
roots in pragmatics.

If this majority view is correct, the question arises how such a division of labor between
semantics and pragmatics could have evolved; why it would be so pervasive across natural
languages; and why it is that some expressions systematically draw from it while others semantically conventionalize. Models of language evolution abound. There are simulation-based models studying
populations of communicating agents
\citep{Hurford1989:Biological-Evol,Steels1995:A-Self-Organizi,LenaertsJansen2005:The-Evolutionar,SteelsBelpaeme2005:Coordinating-Pe,BaronchelliPuglisi2008:Cultural-route-,steels:2011,SpikeStadler2016:Minimal-Require}
and there are mathematical models of language evolution, many coming from game theory
\citep{Warneryd1993:Cheap-Talk-Coor,BlumeKim1993:Evolutionary-St,nowak+krakauer:1999,nowak:2006,Huttegger2007:Evolution-and-t,Skyrms2010:Signals}. Much
of this work has focused on explaining basic properties such as compositionality and
combinatoriality
\citep[e.g.,][]{Batali1998:Computational-S,nowak+krakauer:1999,nowak+etal:2000,KirbyHurford2002:The-Emergence-o,kirby:2002,SmithKirby2003:Iterated-Learni,Gong2007:Language-Evolut,kirby+etal:2015,verhoef+etal:2014,Franke2015:Proto-Syntax},
but little attention has been paid to the interaction between conventional meaning and
pragmatic use. What is more, many mathematical models explain evolved meaning as a regularity
in the overt behavior of agents, abstracting from complex interactions between semantic
representations and pragmatic use. In contrast, we will here look at language users with a
richer linguistic make-up. More precisely, We spell out a model of the co-evolution of conventional meaning and pragmatic reasoning.
The objects of replication and selection are pairs consisting of a set of lexical
meanings and a manner of pragmatic behavior, which together define agents' linguistic choices. Pragmatic behaviors are represented using probabilistic models of pragmatic language use \citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}. Replication
and selection are described by the \emph{replicator mutator dynamic}, a general and established
model of evolutionary change in large and homogeneous populations
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,nowak:2006}. This
approach allows us to study the interaction between (i) pressure towards communicative
efficiency and (ii) infidelity in the transmission of linguistic knowledge, caused by factors
such as inductive learning biases and sparse learning data \citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,KirbyGriffith2014:Iterated-Learni}. Considering the effects that the transmission of
linguistic knowledge through (iterated) learning have is important because neither semantic meanings nor pragmatic usage
patterns are directly observable. Instead, language learners have to infer these unobservables -- lexical meanings and manners of pragmatic behavior --
from the observable behavior in which they result -- overt linguistic choice in a context of interaction. We formalize this process as a form of
Bayesian inference. Our approach thereby contains a well-understood model of
iterated Bayesian learning \citep{griffiths+kalish:2005,griffiths+kalish:2007}, but combines it with functional
selection, here formalized as the most versatile dynamic from evolutionary game theory, the
replicator dynamic \citep{TaylorJonker1978:Evolutionary-St}. Section~\ref{sec:model} introduces
this model.

Section~\ref{sec:si-case-study} applies this model to a case study on scalar implicatures. We
discuss a setting in which the majority view of underspecified lexical meanings and systematic pragmatic
enrichments emerges if selection and transmission infidelity are combined. In particular, we
show that inductive learning biases of Bayesian learners that favor simpler lexical meanings
may play a role in preventing the lexicalization of scalar inferences and thereby lead to the emergence of Gricean-like
pragmatic reasoning types. The results of this case study are critically assessed in the light of
the assumptions that feed our model in Section~\ref{sec:discussion}.

@@

\subsection{Communicative success and learnability}

@@

The idea that language is an adaptation to serve a communicative function is fundamental to
many synchronic and diachronic analyses at least since \citeposs{zipf:1949} explanation of word
frequency rankings as a result of competing hearer and speaker preferences \citep[e.g.,
in][]{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007,
  piantadosi:2014,kirby+etal:2015}. If processes of selection, such as conditional imitation or
reinforcement, favor more communicatively efficient types of behavior, languages are driven
towards semantic expressivity \citep[e.g.,][]{nowak+krakauer:1999,Skyrms2010:Signals}. But
pressure towards communicative efficiency is not the only force that shapes
language. Learnability is another, as natural languages need to be learnable to survive their
faithful transmission across generations. Furthermore, the effects of even small learning biases implicit in acquisition can build up and have quite striking impact on an evolving language in a process of iterated
learning
\citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,kirby+etal:2014}.

While natural languages are pressured for both communicative efficiency and learnability, these forces may
pull in opposite directions \citep[\S7]{christiansen+chater:2008}. Their opposition becomes particularly clear when considering the
extreme (\citealt{kemp+regier:2012,kirby+etal:2015}). A language consisting of a single
form-meaning association is easy to learn but lacking in expressivity. Conversely, a language
that lexicalizes a distinct form for a large number of different meanings is highly expressive
but challenging to acquire.

\subsection{The replicator mutator dynamic}
An elegant formal approach to capture the interaction between communicative efficiency and learnability is
the \emph{replicator mutator dynamic}
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,nowak:2006}. \rvv{The RMD is a population dynamic, which is to say that it describes change in the composition of a community made up of different types of agents. The population vector $\vec{x}$ codifies proportions or frequencies of types in a population, with $x_i$ being the frequency of type $i$. This population is infinite; its carrying capacity is held constant, $\sum\nolimits_i x_i = 1$. This allows us to track change that does not depend on varying population sizes nor their growth rates (but see, e.g., \citealt{nowak:2006} and \citealt[\S 5]{Skyrms2010:Signals} for alternatives that consider, among others, population size and its growth).} In its simplest, discrete-time formulation, the RMD defines the frequency $x'_i$ of each type $i$ in a population at the next time step as a function of: (i) the frequency $x_i$ of each type
$i$ before the update step, (ii) the fitness $f_i$ of each type $i$ before the update, and
(iii) the probability $Q_{ji}$ that an agent who observes the behavior of 
an agent with type $j$ ends up acquiring type $i$:
\begin{align}
  \label{eq:RMD_discrete}
  x'_i = \sum_j Q_{ji} \frac{x_jf_j}{\sum_h x_h f_h}\,.
\end{align}
The RMD consists of two components: fitness-based selection and transmission biases encoded in the so-called mutation matrix $Q$. This
becomes most transparent when we consider an equivalent formulation in terms of a step-wise
application of the discrete-time replicator dynamic \citep{TaylorJonker1978:Evolutionary-St} on the initial population vector $\vec{x}$
and its subsequent multiplication with a mutation matrix $Q$:
\begin{align}
  \label{eq:RMD_discrete_recast}
  x'_i & = (\text{M}(\text{RD}(\vec{x})))_i\,,
\end{align}
where
\begin{align*}
      \left ( \text{RD}(\vec{x}) \right )_i 
         = \frac{x_i f_i}{\sum_k x_k f_k}
 \ \ \ \ \text{and} \ \ \ \ 
  (\text{M}(\vec{x}))_i = (\vec{x} \cdot Q)_i = \left ( \sum_j
          x_j Q_{ji} \right)_i\,.
\end{align*}
%

If the transmission matrix $Q$ is trivial in the sense that $Q_{ji}=1$ whenever $j=i$, the
dynamic reduces to the replicator dynamic. This situation is one where types are always replicated faithfully. The replicator dynamic is a model of fitness-based selection in which the relative frequency of type $i$ will increase with a gradient proportional to its average fitness in the population. In other words, selection comes into play whenever two or more types replicate at different rates due to differences in their fitness. This dynamic is popular and versatile because it can be derived from many abstract processes of biological and cultural
transmission and selection \citep[for overview and several derivations
see][]{Sandholm2010:Population-Game}, including conditional imitation
\citep[e.g.,][]{Helbing1996:A-Stochastic-Be,Schlag1998:Why-Imitate-and} or reinforcement
learning \citep[e.g.,][]{BorgersSarin997:Learning-Throug,Beggs2005:On-the-Converge}. If fitness
$f_i$ is the same for all types $i$, the replicator step is the identity map
$ \left ( \text{RD}(\vec{x}) \right )_i = x_i$. In words, no difference in fitness translates into a lack of functional pressure: the population's type composition stays the same after the replicator step. In such cases the dynamic reduces to a process of iteration of the transmission bias encoded in $Q$: the rate by which one type changes into another after a generational turnover. As detailed below, in this way, the process in (\ref{eq:RMD_discrete}), and its equivalent in (\ref{eq:RMD_discrete_recast}), can contains a model of iterated learning \citep{griffiths+kalish:2005,griffiths+kalish:2007}.

@@

\begin{example}
  Consider a simple and abstract coordination game. Agents are of two types: positive or negative. If
  agents of different type interact with each other, they obtain a payoff of $0$. If negative meets
  negative, each receives a payoff of $1$. If positive meets positive, they get a payoff of $2$.  A population state is completely characterized by the proportion $x$ of negatives. The fitness
  of negatives in population state $x$ is $f_n(x) = x$, that of positives is $f_p(x) = 2-2x$.
  The average fitness is $\Phi(x) = x f_n(x) + (1-x) f_p(x) = 3x^2 - 4x + 2$. Without mutation,
  the replicator dynamic will update $x$ to $\text{RD}(x) = \nicefrac{f(x)x}{\Phi(x)} = \nicefrac{x^2}{\Phi(x)}$. The update
  function $\text{RD}(x)$ of the replicator step is plotted in Fig.~\ref{fig:Updates_RMS} as the
  blue line. Rest points, for which $\text{RD}(x)=x$, are at $x=0$, $x=1$ and $x= \nicefrac{2}{3}$.
  The former are attractors as nearby points converge to them. Points near $x=\nicefrac{2}{3}$
  also move towards 0 or 1. This is schematically pictured in the topmost phase portrait in
  Fig.~\ref{fig:Phase_RD}. Adding mutation changes the dynamic and its rest points. Let us assume that $Q_{ji} = .9$
  when $j=i$. This is the proportion of types that are replicated faithfully. Conversely, a proportion of $.1$ will change their type from positive to negative, and vice-versa. The update effect of mutation on its own are described by $\text{M}(x) = .9 x + .1 (1-x) = .8x + .1$, plotted as the linear green line in Fig.~\ref{fig:Updates_RMS}. As shown in Figure \ref{fig:Phase_RD}, in this example mutation alone has only one stable rest point. It is located at $x = .5$. If we first take the replicator step and
  then the mutation step in sequence, we obtain the replicator mutator dynamic
  $\text{RMD}(x) = \text{M}(\text{RD}(x)) = \nicefrac{.9x^2 - .2x +.2}{3x^2 - 4x + 2}$, which is plotted in red in
  Fig.~\ref{fig:Updates_RMS}. The rest points are at $x=.121$, $x=.903$ and $x=.609$. The
  former two are attractors (see Fig.~\ref{fig:Phase_RD}).
\end{example}

@@

Moving beyond an abstract example, our goal is to apply the RMD to investigate the co-evolution
of lexical representations and pragmatic behavior. To do so, we need to fix three things: (i) what the
relevant types are that populate $x$, (ii) how fitness derives from communicative success and (iii) how the
mutation matrix $Q$ is computed. These issues are addressed, one by one, in the following.

\subsubsection{Types: Lexica and pragmatic strategies}
\label{sec:languages+use}
Types are what evolution operates on. They define an agent's fitness, usually through a payoff
accrued in single interactions with other agents. Often types can be identified as the possible
acts in a game, e.g., either cooperating or defecting in a prisoner's dilemma. In other cases,
they may be thought of as general properties of an agent that influence her fitness, such as
being positive or negative in our previous example (whatever that means). For our present
purposes, types are identified by specific assumptions about their cognitive
make-up. More concretely, since we are interested in the evolutionary competition between different lexical representations and ways of using them in communication, a type is here defined as a pair consisting of a lexicon and a pragmatic strategy of language use.

Agents play signaling games, in which the speaker wants to communicate a world state $s$ with a
message $m$ to a hearer who receives $m$ but does not know $s$
\citep[e.g.][]{lewis:1969,Skyrms2010:Signals}. A lexicon associates each message with a
(possibly fuzzy) set of states. A pragmatic behavior specifies a probabilistic speaker
rule (a probabilistic choice of message for each state) and a probabilistic hearer rule (a
probabilistic choice of state for each message) given a lexicon. There are many ways of making these general notions more concrete. Broadly speaking, there are models of pragmatic language use coming from a game-theoretic tradition \citep[e.g.,][]{benz:2006,benz+vRooij:2007,jaeger:2007b,franke+jaeger:2014} and ones from a Bayesian tradition \citep[e.g.,][]{frank+goodman:2012,goodman+stuhlmueller:2013,GoodmanFrank2016:Pragmatic-Langu}. We will not focus on the particular design choices that make up each approach \citep{qing+franke:2015}. Instead, we will stress how the approach, common to all of these models, of pragmatic language use as {\em rational language use} can be embedded within the replicator mutator dynamic to answer evolutionary questions at the semantics-pragmatics interface. Here is what we will assume in the remainder of this paper.\footnote{Readers more familiar with the Bayesian tradition should note that the definitions below follow the game-theoretic tradition in assuming rational sender choice to consist in sending messages that maximize expected utility, rather than message informativity (see \citealt{qing+franke:2015} for detailed discussion).}

@@

In words, an agent's linguistic behavior -- what message to send when / which message to interpret how -- is defined by (i) her reasoning level and (ii) her lexicon. In what follows, this is what we identify an agent's type with in order to analyze under which conditions combinations of (i) and (ii) emerge through cultural evolution. Intuitively, literal behavior (level-$0$ reasoning) results from unreflected language use. Such agents produce and comprehend solely based on the truth-conditions of their lexicon. In a Gricean spirit, pragmatic behavior (level-$n+1$ reasoning) results from mutual reasoning about rational language use. Such agents produce and comprehend by reasoning about their interlocutor's likely behavior interpretation/production patterns. This reasoning is based on a conjecture resulting from their own behavior if they were in the role of their interlocutor. Higher order behavior is therefore, even if in tendency rational, quite simple. Crucially, we do not assume that agents know each other's types when interacting (e.g., what lexicon the interlocutor is using).

The following examples illustrate these behaviors numerically. A literal interpreter with lexicon $\Lbound$ assigns $\ssome$

@@

Recall that types are a combination of a lexicon and a manner of language use. We analyze the
model's predictions in populations of types with one of the two behaviors introduced earlier;
literal or pragmatic. The former correspond to level $0$ reasoners and the latter to level
$1$. Accordingly, we consider a total of $432$ types. Six are variants of pragmatic language
users with $\Llack$-like lexica. We refer to these as \emph{target types}, because they
represent lexica and language use that conform to the majority view of scalar
implicatures. Twelve types are either literal or pragmatic speakers with lexica of the
$\Lbound$ kind. We refer to these as \emph{competitor types}, because they are expected to be
the target types' main contenders in evolutionary competition. Finally, note that while
different types may lexicalize the same representations, they may nevertheless map different
states to different overt messages. More informally, they speak different languages that
lexicalize the same concepts. Consequently, more often than not different types of the same
kind fail to understand each other.

@@

\hl{The discussion on derivation costs is not very satisfying on my opinion. For instance, the condition of "simple representations being favored over more complex ones" is compatible with many more cost specifications.}
